---
title: "Prepare Silva 138.2 SSU taxonomy data for DADA2"
author: "Michael McLaren"
date: "2021-08-02"
output:
  html_document:
    toc: false
    toc_float: true
    self_contained: true
    highlight: tango
---

```{r, include = FALSE}
# knitr chunk options
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE
)
```

This script downloads Silva SSU 138.2 taxonomic database files and formats them for use with DADA2. 

```{r}
library(dplyr) 
library(tibble) 
library(stringr) 
library(purrr) 

packageVersion("dada2")

silva_version <- "138.2"

# Path for storing precursor files and output
path <- here::here()
```

Create the locations where we will download the precursor files and save our output,
```{r}
dir.create(file.path("precursors"))
dir.create(file.path(path, "output"))
```

The precursor files are available from Silva at

- https://www.arb-silva.de/no_cache/download/archive/release_138.1/Exports
- https://www.arb-silva.de/no_cache/download/archive/release_138_1/Exports

These two listings seem to be of identical files.

Download the precursor files and their MD$ checksums,
```{r}
base_url <- str_glue("https://www.arb-silva.de/fileadmin/silva_databases/release_{silva_version}")

fns <- c(
  ref = "SILVA_{silva_version}_SSURef_tax_silva.fasta.gz",
  ref_nr = "SILVA_{silva_version}_SSURef_NR99_tax_silva.fasta.gz",
  tax = "tax_slv_ssu_{silva_version}.txt.gz"
) %>%
  map_chr(str_glue)
ftb <- fns %>%
  enframe("type", "file") %>%
  mutate(
    url = case_when(
      type == "tax" ~ file.path(base_url, "Exports", "taxonomy", file),
      TRUE ~ file.path(base_url, "Exports", file)
    ),
    destfile = file.path(path, "precursors", file),
  ) %>%
  mutate(
    across(-type, str_c, ".md5", .names = "{.col}_md5"),
    file_exists = file.exists(destfile)
  )
# Only download the main files if they haven't already been downloaded
ftb %>% 
  filter(!file_exists) %>%
  select(url, destfile) %>% 
  pwalk(download.file, method = "wget")
ftb %>% select(url = url_md5, destfile = destfile_md5) %>% pwalk(download.file)
```

Check files against their MD5s,
```{r}
ftb <- ftb %>%
  mutate(
    local_md5 = map_chr(destfile, tools::md5sum),
    remote_md5 = map(destfile_md5, read.table) %>% map_chr(pull, 1)
  )
stopifnot(all.equal(ftb$local_md5, ftb$remote_md5))
```

Extract the `tax_slv_ssu_{silva_version}.txt.gz` file,
```{r}
precursors <- ftb %>% select(type, destfile) %>% deframe
system2("gunzip", args = precursors["tax"])
precursors["tax"] <- precursors["tax"] %>% str_extract(".+[^\\.gz$]")
```

Download the Silva license,
```{r}
download.file(
  file.path(base_url, "LICENSE.txt"),
  file.path(path, "SILVA_LICENSE.txt")
)
```

```{r}
list.files(file.path(path, "precursors"))
```

Create the DADA2-formatted training files, with and without the species level, from the NR99 reference set,
```{r}

# Locations to save the files
output <- c(
  train = "silva_nr99_v{silva_version}_train_set.fa.gz",
  train_with_species = "silva_nr99_v{silva_version}_wSpecies_train_set.fa.gz",
  species = "silva_species_assignment_v{silva_version}.fa.gz"
) %>%
  map_chr(str_glue) %>%
  map_chr(~file.path(path, "output", .))

dada2:::makeTaxonomyFasta_SilvaNR(
  precursors["ref_nr"],
  precursors["tax"],
  output["train"]
)
dada2:::makeTaxonomyFasta_SilvaNR(
  precursors["ref_nr"],
  precursors["tax"],
  output["train_with_species"],
  include.species = TRUE
)
```
The addition of "99" in "nr99" distinguishes these files from the previous versions that were derived from the mothur wiki files.

Create the DADA2-formatted species database file using the full reference set,
```{r}
dada2:::makeSpeciesFasta_Silva(
  precursors["ref"],
  output["species"]
)
```

```{r}
list.files(file.path(path, output))
```

## Session info

```{r, R.options = list(width = 83)}
sessioninfo::session_info()
```
